Learning to Pivot with Adversarial Networks

### 2017-02-01 tex/icml2017/paper.tex
save

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2016}

@@

Many inference problems involve data generation processes that are not uniquely
specified or are uncertain in some way. In a scientific context, the presence of
several plausible data generation processes is often associated to the presence
of systematic uncertainties. Robust inference is possible if it is based on a
pivot -- a quantity whose distribution is invariant to the unknown value of the
(categorical or continuous) nuisance parameters that parametrizes this family of
generation processes. In this work,  we introduce a flexible training procedure
based on adversarial networks for enforcing the pivotal property on a predictive
model. We derive theoretical results showing that the proposed algorithm tends
towards a minimax solution corresponding to a predictive model that is both
optimal and independent of the nuisance parameters (if that model exists) or
for which one can tune the trade-off between power and robustness. Finally, we
demonstrate the effectiveness of this approach with a toy example and an example
from particle physics.

@@

Assuming a probability model $p(X,Y,Z)$, where $X$ are the data, $Y$ are the
target labels, and $Z$ are the nuisance parameters, we consider the problem of
learning a predictive model for $Y$ conditional on the observed values of $X$
that is robust to uncertainty in the unknown value of $Z$. More specifically, we
introduce a learning procedure for enforcing the pivotal property
\citep{degroot1986probability}  on a
predictive model by jointly training two neural networks in an adversarial
fashion. We derive theoretical results proving that the proposed procedure
converges towards a predictive model that is both optimal and independent of the
nuisance parameters (if that models exists) or for which one can tune the
trade-off between power and robustness. Finally, we demonstrate the
effectiveness of this approach with a toy example and an example from particle physics.

@@

We augment our initial objective so that inference based on $f(X ; \theta_f)$ will be
robust to the value $z \in {\cal Z}$ of the nuisance parameter $Z$  -- which remains unknown at
test time. A formal way of enforcing robustness is to require that the distribution of
$f(X ; \theta_f)$ conditional on $Z$ (and possibly $Y$) be invariant with
 the nuisance parameter $Z$. Thus, we wish to find a function $f$ such that
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.
In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameters. This implies that  $f(X; \theta_f)$ and $Z$ are independent random variables.

@@

In this work, we repurpose adversarial networks as a means to constrain the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Fig.~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $f(X; \theta_f)$ and produces as output a function
modeling the posterior probability density $p_{\theta_r}(z | f(X;\theta_f)=s)$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

@@

Despite the theoretical characterization of the minimax solution of Eqn.~\ref{eqn:min_thetaf}
that motivates this work,
Algorithm~\ref{alg:adversarial-training} remains a heuristic in the sense
that we do not provide formal guarantees of convergence of the procedure towards
that solution. Empirical evidence gathered here and in similar minimax-based
 works (see e.g., \citep{goodfellow2014generative,EdwardsS15}) suggests
that the alternating optimization algorithm works sufficiently well in practice,
but further research studying its stability and convergence would certainly
be beneficial for a large body of works.

>>>>

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage[accepted]{icml2016}

@@

Many inference problems involve data generation processes that are not uniquely
specified or are uncertain in some way. In a scientific context, the presence of
several plausible data generation processes is often associated to the presence
of systematic uncertainties. Robust inference is possible if it is based on a
pivot -- a quantity whose distribution does not depend on the unknown value of the
(categorical or continuous) nuisance parameters that parametrizes this family of
generation processes. In this work,  we introduce a flexible training procedure
based on adversarial networks for enforcing the pivotal property, i.e., the stastistical independence of a predictive
model with respect to nuisance parameters. We derive theoretical results showing that the proposed algorithm tends
towards a minimax solution corresponding to a predictive model that is both
optimal and independent of the nuisance parameters (if that model exists) or
for which one can tune a trade-off between power and robustness. Finally, we
demonstrate the effectiveness of this approach with a toy example and an example
from particle physics.

@@

Assuming a probability model $p(X,Y,Z)$, where $X$ are the data, $Y$ are the
target labels, and $Z$ are the nuisance parameters, we consider the problem of
learning a predictive model $f(X)$ for $Y$ conditional on the observed values of $X$
that is robust to uncertainty in the unknown value of $Z$. More specifically, we
introduce a learning procedure based on adversarial networks
for enforcing that  $f(X)$ is a pivot, i.e., a quantity whose distribution
does not depend on $Z$.
We derive theoretical results proving that the proposed procedure
converges towards a predictive model that is both optimal and statistically independent of the
nuisance parameters (if that model exists) or for which one can tune a
trade-off between power and robustness, e.g., as driven by the requirements of a higher level statistical analysis.
Finally, we demonstrate the
effectiveness of this approach with a toy example and an example from particle physics.

@@

We augment our initial objective so that inference based on $f(X ; \theta_f)$ will be
robust to the value $z \in {\cal Z}$ of the nuisance parameter $Z$  -- which remains unknown at
test time. A formal way of enforcing robustness is to require that the distribution of
$f(X ; \theta_f)$ conditional on $Z$ (and possibly $Y$) be invariant with
 the nuisance parameter $Z$. Thus, we wish to find a function $f$ such that
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.
In words, we are looking for a predictive function $f$
which is a pivotal quantity (see e.g., \citep{degroot1986probability}) with respect to the
nuisance parameters. This implies that  $f(X; \theta_f)$ and $Z$ are independent random variables.

@@

In this work, we repurpose adversarial networks as a means to constrain the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Fig.~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$ and produces as output a function
modeling the posterior probability density $p_{\theta_r}(z | f(X;\theta_f)=s)$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

@@

Despite the theoretical characterization of the minimax solution of Eqn.~\ref{eqn:min_thetaf}
that motivates this work,
Algorithm~\ref{alg:adversarial-training} remains a heuristic in the sense
that we do not provide formal guarantees of convergence of the procedure towards
that solution. Empirical evidence gathered here and in similar minimax-based
 works suggests
that the alternating optimization algorithm works sufficiently well in practice,
but further research studying its stability and convergence would certainly
be beneficial for a large body of works.
